---
title: "IFP Pie Competition 2025 Results"
output: 
  html_document:
    css: "custom_style.css"
---

# 2025 Pie of the Year
Pie of the Year -- the people's choice -- was decided by ranked choice voting with an instant runoff, where the win condition was a majority of all active ballots. Cast Iron Apple Pie took the lead early in the first round, but it became a nail biter as Australian Vegan Meat Pie nearly caught up in round 3. Nevertheless, the *Cast Iron Apple Pie* prevailed in the third round (same number of fire alarms last night?) and ...  
\
*Beez Africa* reigns as Pie Champion 2025, with all of the glory and eternal bragging rights that the title bestows. She joins [Pie Champion 2024 Amy Nice](PieContestWinner24.html) in the IFP Hall of Pie Champions.
\
\
```{r, echo=FALSE, warning=FALSE, message=FALSE}

library(tidyverse)
library(googlesheets4)
library(ggplot2)
library(kableExtra)

# Authenticate Google Sheets (only needed the first time)
gs4_deauth()
knitr::opts_chunk$set(echo = TRUE)
sheet_url <- "https://docs.google.com/spreadsheets/d/1ZdX-om0XlLlOnRoGc7Iv2i5eRie2es9Zn7ECK4JEWX4/edit?resourcekey=&gid=1594056022#gid=1594056022"

data <- read_sheet(sheet_url)

# Extract columns containing rankings
ranking_columns <- data %>%
  select(starts_with("Please rank")) %>%
  rename_with(~ paste0("Pie_", seq_along(.)))
# Convert rankings to numeric
convert_rankings <- function(x) {
  as.numeric(str_replace(x, "th|st|nd|rd", ""))
}

ranked_ballots <- ranking_columns %>%
  mutate(across(everything(), convert_rankings))

run_rcv <- function(ballots) {
  total_ballots <- nrow(ballots)  # Total ballots cast
  rounds <- list()  # Store each round's results
  
  while (TRUE) {
    # Count first-choice votes
    first_choices <- apply(ballots, 1, function(row) {
      if (all(is.na(row))) {
        return(NA)  # Ballot is exhausted
      } else {
        return(names(row)[which.min(row)])
      }
    })
    
    first_choice_table <- table(first_choices, useNA = "ifany")
    
    # Convert to data frame for easier processing
    round_result <- as.data.frame(first_choice_table, stringsAsFactors = FALSE) %>%
      rename(Candidate = first_choices, Votes = Freq)
    
    # Calculate number of exhausted ballots
    exhausted_votes <- sum(is.na(first_choices))
    round_result <- round_result %>% 
      mutate(Candidate = if_else(is.na(Candidate), "Exhausted", Candidate))
      
    
    rounds <- append(rounds, list(round_result))
    
  # Check for majority of total ballots cast
valid_votes <- round_result$Votes[round_result$Candidate != "Exhausted" & !is.na(round_result$Votes)]
active_total <- sum(valid_votes)
majority_threshold <- active_total / 2
if (any(valid_votes > majority_threshold)) {
  winner <- round_result$Candidate[which.max(round_result$Votes)]
  return(list(winner = winner, rounds = rounds))
}

    
    # Eliminate candidate(s) with the fewest votes, including those with zero votes
    active_candidates <- round_result %>%
      filter(Candidate != "Exhausted")
    min_votes <- min(active_candidates$Votes)
    eliminated <- active_candidates$Candidate[active_candidates$Votes == min_votes]
    
    # Update ballots by marking eliminated candidates as NA
    ballots[colnames(ballots) %in% eliminated] <- NA

    
    # Recalculate rankings for each ballot
   ballots <- t(apply(ballots, 1, function(row) {
  # Re-rank only non-NA values
  rank <- rank(row, na.last = "keep", ties.method = "min")
  ifelse(is.na(row), NA, rank)  # Keep NA values for eliminated candidates
}))

# Convert back to data frame
ballots <- as.data.frame(ballots)

    # Terminate if all ballots are exhausted
    if (total_ballots == exhausted_votes) {
      return(list(winner = "No winner (all ballots exhausted)", rounds = rounds))
    }
  }
}


# Run RCV
rcv_results <- run_rcv(ranked_ballots)

# Prepare Data for Plotting
all_pies <- c(paste0("Pie_", 1:8), "Exhausted")


plot_data <- bind_rows(lapply(seq_along(rcv_results$rounds), function(round) {
  round_df <- rcv_results$rounds[[round]]
  colnames(round_df) <- c("Candidate", "Votes")

  # Add missing pies with zero votes
  complete_df <- data.frame(
    Candidate = all_pies,
    Round = paste("Round", round),
    stringsAsFactors = FALSE
  )
  
  # Combine actual data with the complete set
  complete_df <- left_join(complete_df, round_df, by = c("Candidate"))
  
  # Fill in actual votes and ensure zero for missing pies
  complete_df <- complete_df %>%
    mutate(Votes = ifelse(is.na(Votes), 0, Votes)) %>%
    select(Candidate, Votes, Round)
  
  complete_df
}))

library(ggplot2)
library(dplyr)
library(forcats)

# Preprocess plot_data
plot_data <- plot_data %>%
  mutate(
    # Replace underscores with spaces and fix naming
    Candidate = ifelse(Candidate == "Exhausted", "Exhausted", gsub("_", " ", Candidate)),
    # Ensure consistent ordering with Exhausted last
    Candidate = factor(Candidate, levels = c((paste("Pie", 1:8)), "Exhausted"), ordered = TRUE)
  ) %>%
  group_by(Round) %>%
  mutate(Votes = Votes) %>%
  ungroup()

# Define IFP palette
ifp_palette <- c(
  "Pie 1" = "#ff6565",
  "Pie 2" = "#ff9762",
  "Pie 3" = "#3368ce",
  "Pie 4" = "#39d794",
  "Pie 5" = "#41c4fc",
  "Pie 6" = "#fbdb36",
  "Pie 7" = "#ff9ee3",
  "Pie 8" = "#ff656580",
  "Pie 9" = "#3368ce80",
  "Pie 10" = "#fbdb3680",
  "Exhausted" = "#d3d3d3"
)



# Plot RCV results with IFP palette
ggplot(plot_data, aes(x = Votes, y = fct_relevel(Candidate, paste("Pie", 1:8), "Exhausted"), fill = Candidate)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ Round, scales = "free_x", ncol = 2) +
  labs(
    title = "Ranked Choice Voting Results",
    x = "Number of Total Ballots",
    y = NULL,  # Remove y-axis label
    fill = "Candidate"
  ) +
  scale_y_discrete(limits = c("Exhausted", paste("Pie", 8:1))) +
  scale_x_continuous(
    limits = c(0, 12),
    breaks = seq(0, 12, by = 2),  
    labels = seq(0, 12, by = 2) 
  ) +
  scale_fill_manual(values = ifp_palette) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18, face = "bold", color = "#373737"),
    axis.text.x = element_text(size = 12, color = "#373737"),  # Adjust x-axis text
    strip.background = element_rect(fill = "#b17ada", color = NA),
    strip.text = element_text(color = "white", face = "bold"),
    panel.background = element_rect(fill = "#fcfbeb", color = NA),  # Cream background
    plot.background = element_rect(fill = "#fcfbeb", color = NA),   # Cream background
    legend.background = element_rect(fill = "#fcfbeb", color = NA)  # Cream legend background
  ) 
```

# Criteria Ratings
Our discerning judges also provided their opinions on a couple criteria, handing out anywhere from 1 to 5 stars.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Ensure consistent data types
non_rcv_metrics <- data %>%
  select(starts_with("Pie")) %>%
  mutate(across(everything(), as.character))  # Convert all columns to character

pies <- tribble(
  ~Pie, ~PieName,
  "Pie 1", "Pie 1: Sweet Potato Pecan Pie",
  "Pie 2", "Pie 2: Chocolate Pecan Pie",
  "Pie 3", "Pie 3: Coconut Sweet Potato",
  "Pie 4", "Pie 4: Cast Iron Apple Pie",
  "Pie 5", "Pie 5: Ma's Chocolate Pie",
  "Pie 6", "Pie 6: Chocolate Silk Pie",
  "Pie 7", "Pie 7: Vegan Pumpkin Pie",
  "Pie 8", "Pie 8: Australian Vegan Meat Pie"
)


# Restructure data to ensure rows are pies and columns are criteria
average_scores <- non_rcv_metrics %>%
  pivot_longer(cols = everything(), names_to = "Metric", values_to = "Score") %>%
  # Convert "5 stars" to numeric 5 and handle other cases
  mutate(
    Score = case_when(
      Score == "5 stars" ~ 5,
      TRUE ~ as.numeric(Score)
    )
  ) %>%
  # Separate "Pie" and the criterion (e.g., Taste, Memorability)
  separate(Metric, into = c("Pie", "Criterion"), sep = "\\s\\[") %>%
  mutate(
    Pie = str_trim(Pie),  # Remove any leading/trailing whitespace
    Criterion = str_remove(Criterion, "\\]")  # Remove closing bracket
  ) %>%
  left_join(pies, by="Pie") %>% 
  select(-Pie) %>%
  rename(Pie = PieName) %>%
  # Calculate average scores grouped by Pie and Criterion
  group_by(Pie, Criterion) %>%
  summarise(Average = round(mean(Score, na.rm = TRUE), 2), .groups = "drop") %>%
  # Pivot data to make rows as pies and columns as criteria
  pivot_wider(names_from = Criterion, values_from = Average) %>%
  select(1,5,4,3,2) %>% 
  mutate(number = as.numeric(sub("Pie (\\d+):.*", "\\1", Pie)),
         Pie = factor(Pie, levels = Pie[order(number)], ordered = TRUE)) %>%
  arrange(number)


# Restructure data to ensure rows are pies and columns are criteria
variance_scores <- non_rcv_metrics %>%
  pivot_longer(cols = everything(), names_to = "Metric", values_to = "Score") %>%
  # Convert "5 stars" to numeric 5 and handle other cases
  mutate(
    Score = case_when(
      Score == "5 stars" ~ 5,
      TRUE ~ as.numeric(Score)
    )
  ) %>%
  # Separate "Pie" and the criterion (e.g., Taste, Memorability)
  separate(Metric, into = c("Pie", "Criterion"), sep = "\\s\\[") %>%
  mutate(
    Pie = str_trim(Pie),  # Remove any leading/trailing whitespace
    Criterion = str_remove(Criterion, "\\]")  # Remove closing bracket
  ) %>%
  left_join(pies, by="Pie") %>% 
  select(-Pie) %>%
  rename(Pie = PieName) %>%
  # Calculate average scores grouped by Pie and Criterion
  group_by(Pie, Criterion) %>%
  summarise(Average = round(var(Score, na.rm = TRUE), 2), .groups = "drop") %>%
  # Pivot data to make rows as pies and columns as criteria
  pivot_wider(names_from = Criterion, values_from = Average) %>%
  select(1,5,4,3,2) %>% 
  mutate(number = as.numeric(sub("Pie (\\d+):.*", "\\1", Pie)),
         Pie = factor(Pie, levels = Pie[order(number)], ordered = TRUE)) %>%
  arrange(number)

# Display averages in a styled table
library(kableExtra)
library(DT)

datatable(
  average_scores,
  caption = "Average Stars for Non-RCV Metrics",
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    lengthChange = FALSE, # Disable the option to change the number of rows
    searching = FALSE,     # Optional: Disable search functionality
    info = FALSE,
    paging = FALSE,
    columnDefs=list(list(orderData=5,targets=0),
                    list(visible=FALSE,targets=5))
  ),
  rownames = FALSE
)




datatable(
  variance_scores,
  caption = "Variance in Stars for Non-RCV Metrics",
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    lengthChange = FALSE, # Disable the option to change the number of rows
    searching = FALSE,     # Optional: Disable search functionality
    info = FALSE,
    paging = FALSE,
    columnDefs=list(list(orderData=5,targets=0),
                    list(visible=FALSE,targets=5))
  ),
  rownames = FALSE
)


```


